Appendix 1: Quantifying Drought

### 2021-01-04 analysis/supplementary-materials/appendix_SPEI.Rmd
add imputed data to table

xa_spei <-
  xa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(latlon) %>%
  group_nest() %>% 
  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(xa_spei, data)) %>% 
  select(-cb)
```
# EMBRAPA stations

@@

@@

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

@@

```{r}
spei_table <-
  comb %>% 
  group_by(year, dataset) %>% 
  mutate(min = min(spei)) %>% 
  filter(spei == min) %>% 
  mutate(month = month(yearmonth, label = TRUE)) %>% 
  # summarize(min = min, month = month) %>%
  filter(year %in% c(2002, 2004, 2006, 2009, 1997, 2005, 2010)) %>% 
  select(year, dataset, spei, month) %>% 
  mutate(dataset = str_remove(dataset, "_spei")) %>% 
  pivot_wider(names_from = dataset, values_from = c(spei, month)) %>% 
  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), .after = "year") %>% 
  mutate(`Global SPEI` = glue("{round(spei_g, 2)} ({month_g})"),
         `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
         `EMBRAPA` = glue("{round(spei_st, 2)} ({month_st})"),
         `TRMM` = glue("{round(spei_trmm, 2)} ({month_trmm})")) %>% 
  ungroup()
```

```{r}
library(gt)
gt(spei_table) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#ffffb2")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < 0 & spei_g >= -1
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < 0 & spei_st >= -1
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < 0 & spei_xa >= -1
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < 0 & spei_trmm >= -1
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#fecc5c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -1 & spei_g >= -1.5
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -1 & spei_st >= -1.5
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -1 & spei_xa >= -1.5
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -1 & spei_trmm >= -1.5
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#fd8d3c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -1.5 & spei_g >= -2
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -1.5 & spei_st >= -2
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -1.5 & spei_xa >= -2
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -1.5 & spei_trmm >= -2
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#e31a1c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -2
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -2
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -2
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -2
      )
    )
  ) %>% 
  cols_hide(starts_with(c("spei_", "month_"))) %>% 
  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM))
```

@@

```{r}
spei_plot <- 
  comb %>% 
  group_by(year) %>% 
  mutate(min_mon = spei == min(spei, na.rm = TRUE)) %>% 
  mutate(drought_label = case_when(
    year %in% c(2004, 2006) & min_mon ~ "weak ENSO",
    year %in% c(2002, 2009) & min_mon ~ "moderate ENSO",
    year == 1997 & min_mon ~ "severe ENSO",
    year %in% c(2005, 2010) & min_mon ~ "severe SST"
    )) %>% 
  mutate(dataset = case_when(
    dataset == "g_spei"  ~ "Global SPEI",   
    dataset == "xa_spei" ~ "Xavier et al.",
    dataset == "st_spei" ~ "EMBRAPA",
    dataset == "trmm_spei" ~ "TRMM"
  ))
```

@@

p <-
  ggplot(spei_plot, aes(x = yearmonth, y = spei)) +
  geom_line(aes(color = dataset), size = 0.5) +
  geom_segment(data = labs,
               aes(xend = yearmonth, yend = spei -1)) +
  geom_label(data = labs,
             aes(x = yearmonth, y = spei-1, label = year, fill = drought_label)) +
  scale_x_yearmonth(date_breaks = "2 years") +
  scale_color_brewer(type = "qual", palette = 2) +
  coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
  theme_bw()
annotate_spei(p)
ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
```

>>>>

xa_spei <-
  xa_mon %>% 
  mutate(cb = precip - eto) %>% 
  group_by(latlon) %>%
  group_nest() %>% 
  mutate(xa_spei = map(data, ~as.numeric(spei(.x$cb, scale = 3)$fitted))) %>% 
  unnest(c(xa_spei, data)) %>% 
  select(-cb)
```

# EMBRAPA stations

@@

# Imputed data (averaged across for BDFFP)

```{r}
bdffp_spei <- 
  bdffp %>% group_by(date) %>% 
  summarize(bdffp_spei = mean(spei)) %>% 
  mutate(yearmonth = yearmonth(date), date = NULL)
```

@@

comb <-
  full_join(xa_df, gspei) %>% 
  full_join(trmm_df) %>% 
  full_join(embrapa_wide %>% select(yearmonth, st_spei)) %>% 
  full_join(bdffp_spei) %>% 
  pivot_longer(-yearmonth, names_to = "dataset", values_to = "spei") %>% 
  filter(yearmonth > yearmonth(ymd("1990-01-01"))) %>% 
  mutate(year = year(yearmonth), .before = yearmonth)
```

@@

```{r}
spei_table <-
  comb %>% 
  group_by(year, dataset) %>% 
  mutate(min = min(spei, na.rm = TRUE)) %>% 
  filter(spei == min) %>% 
  mutate(month = month(yearmonth, label = TRUE)) %>% 
  # summarize(min = min, month = month) %>%
  filter(year %in% c(2002, 2004, 2006, 2009, 1997, 2005, 2010)) %>% 
  select(year, dataset, spei, month) %>% 
  mutate(dataset = str_remove(dataset, "_spei")) %>% 
  pivot_wider(names_from = dataset, values_from = c(spei, month)) %>% 
  add_column(cause = c("ENSO", "ENSO", "ENSO", "AMO", "ENSO", "ENSO", "AMO"), .after = "year") %>% 
  mutate(`Global SPEI` = glue("{round(spei_g, 2)} ({month_g})"),
         `Xavier et al.` = glue("{round(spei_xa, 2)} ({month_xa})"),
         EMBRAPA = glue("{round(spei_st, 2)} ({month_st})"),
         TRMM = glue("{round(spei_trmm, 2)} ({month_trmm})"),
         BDFFP = glue("{round(spei_bdffp, 2)} ({month_bdffp})")) %>% 
  ungroup()
```

```{r}
library(gt)
table <- 
  gt(spei_table) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#ffffb2")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < 0 & spei_g >= -1
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < 0 & spei_st >= -1
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < 0 & spei_xa >= -1
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < 0 & spei_trmm >= -1
      ),
      cells_body(
        columns = vars(BDFFP),
        rows = spei_bdffp < 0 & spei_bdffp >= -1
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#fecc5c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -1 & spei_g >= -1.5
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -1 & spei_st >= -1.5
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -1 & spei_xa >= -1.5
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -1 & spei_trmm >= -1.5
      ),
      cells_body(
        columns = vars(BDFFP),
        rows = spei_bdffp < -1 & spei_bdffp >= -1.5
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#fd8d3c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -1.5 & spei_g >= -2
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -1.5 & spei_st >= -2
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -1.5 & spei_xa >= -2
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -1.5 & spei_trmm >= -2
      ),
      cells_body(
        columns = vars(BDFFP),
        rows = spei_bdffp < -1.5 & spei_bdffp >= -2
      )
    )
  ) %>% 
  tab_style(
    style = list(
      cell_fill(color = "#e31a1c")
    ),
    locations = list(
      cells_body(
        columns = vars(`Global SPEI`),
        rows = spei_g < -2
      ),
      cells_body(
        columns = vars(EMBRAPA),
        rows = spei_st < -2
      ),
      cells_body(
        columns = vars(`Xavier et al.`),
        rows = spei_xa < -2
      ),
      cells_body(
        columns = vars(TRMM),
        rows = spei_trmm < -2
      ),
      cells_body(
        columns = vars(BDFFP),
        rows = spei_bdffp < -2
      )
    )
  ) %>% 
  cols_hide(starts_with(c("spei_", "month_"))) %>% 
  tab_spanner("Lowest SPEI for the year (Month)", vars(`Global SPEI`, EMBRAPA, `Xavier et al.`, TRMM, BDFFP))
table
gtsave(table, here("analysis", "supplementary-materials", "spei_table.png"))
```

@@

```{r}
spei_plot <- 
  comb %>% 
  group_by(year) %>% 
  mutate(min_mon = spei == min(spei, na.rm = TRUE)) %>% 
  mutate(drought_label = case_when(
    year %in% c(2004, 2006) & min_mon ~ "weak ENSO",
    year %in% c(2002, 2009) & min_mon ~ "moderate ENSO",
    year == 1997 & min_mon ~ "severe ENSO",
    year %in% c(2005, 2010) & min_mon ~ "severe SST"
    )) %>% 
  mutate(dataset = case_when(
    dataset == "g_spei"  ~ "Global SPEI",   
    dataset == "xa_spei" ~ "Xavier et al.",
    dataset == "st_spei" ~ "EMBRAPA",
    dataset == "trmm_spei" ~ "TRMM",
    dataset == "bdffp_spei" ~ "BDFFP imputed"
  ))
```

@@

p <-
  ggplot(spei_plot, aes(x = yearmonth, y = spei)) +
  geom_line(aes(color = dataset), size = 0.5) +
  geom_segment(data = labs,
               aes(xend = yearmonth, yend = spei -1)) +
  geom_label(data = labs,
             aes(x = yearmonth, y = spei-1, label = year, fill = drought_label)) +
  scale_x_yearmonth(date_breaks = "2 years") +
  scale_color_brewer(type = "qual", palette = "Set1") +
  coord_cartesian(xlim = c(yearmonth(ymd("1997-01-01")), yearmonth(ymd("2010-11-01")))) +
  theme_bw()
annotate_spei(p, alpha = 0.7)
ggsave(here("analysis", "supplementary-materials", "spei_comparison.png"))
```
